---
title: "Auckland Cycle analysis"
author: "Lawrence May"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Total number of cyclists counted for each day, with rainfall.

```{r}
library(tidyverse)

#Read in bike data
bike_files<- list.files("/Users/lawrence/Google Drive/UNI/current/Stats 369/a1/",pattern=".csv",full=TRUE)
bike_data<-map(bike_files,read_csv)

#Join into one dataframe
bike_data <- dplyr::bind_rows(bike_data[[1]],bike_data[[2]],bike_data[[3]])

bike <- subset(bike_data, select = -c(Date))
bike[is.na(bike)] <- 0

#Take sum of all cyclists on each day
bike<-rowSums(bike)
bike_data <- data.frame(bike_data$Date,bike)
bike_data %>% rename(Date = bike_data.Date, Count = bike) -> bikes

#Read in rain data
rain_files<- list.files("/Users/lawrence/Google Drive/UNI/current/Stats 369/a1/",pattern=".txt",full=TRUE)
rain_data<-map(rain_files,read_csv,skip=9)

#Join into one dataframe
rain <- dplyr::bind_rows(rain_data[[1]],rain_data[[2]])
rain %>% filter(Station == 22719) -> rain1
rain %>% filter(Station == 37852) -> rain2

#Aggregate daily rainfall for each locationn
aggregate(rain1$`Amount(mm)` ~ rain1$`Date(NZST)`, data=rain1, FUN=sum)  -> rain1
aggregate(rain2$`Amount(mm)` ~ rain2$`Date(NZST)`, data=rain2, FUN=sum)  -> rain2

colnames(rain1) <- c("Date", "Rain")
colnames(rain2) <- c("Date", "Rain")
rain <- rbind(rain1,rain2)

#Take average of daily rain in Mangere and Albany to get approximation for average rainfall across Auckland
aggregate(Rain ~ Date, data=rain, FUN=mean)  -> rain

bikes %>% separate(col=Date, into=c("dow","day","month","year")) %>% unite(Date,c(day,month,year), sep = " ",remove = FALSE) %>% select(-c(day)) -> bikes

#Change date to same format and join on date
bikes$Date <- as.Date(bikes$Date, format='%d %b %Y')
rain$Date <- as.Date(as.character(rain$Date), format='%Y%m%d')
combined_df <- inner_join(bikes,rain,by = 'Date')
head(combined_df)
```

Some graphical analysis

```{r}
#Create season factor variable
spring <- c("Sep", "Oct", "Nov")
summer <- c("Dec", "Jan", "Feb")
autumn <- c("Mar", "Apr", "May")
winter <- c("Jun", "Jul", "Aug")

#Insert into dataframe 
combined_df <- mutate(combined_df, season = factor(case_when(month %in% spring ~ "spring",
                                             month %in% summer ~ "summer", month %in% autumn ~ "autumn",
                                             month %in% winter ~ "winter",
                                             TRUE                ~ NA_character_)))

ggplot(combined_df, aes(Date, Count, color = season)) + geom_point()+ggtitle("Number of cyclists over time and by season")
```
There appears to be an increasing trend of the number of cyclists over time, with a strong seasonal component (higher number of cyclist in spring and summer).


```{r}
ggplot(combined_df, aes(season, Count, color = Rain)) + geom_boxplot()+ggtitle("Number of cyclists by season and rain") + geom_point()
```

There is a strong seasonal pattern with most cyclists in the summer, spring and autumn months and far less in the winter months. Rain does appear to have a negative effect as well, and seems to be able to explain many of the negative outliers (light blue = heavy rain).


```{r}
combined_df$dow<-factor(combined_df$dow , levels=c('Mon','Tue','Wed','Thu','Fri','Sat','Sun'))
ggplot(combined_df, aes(dow, Count, color = Rain)) + geom_boxplot()+ggtitle("Number of cyclists by day of the week and rain")+ geom_point()
```
There also appears to be a Day of the week effect, with most cyclists riding on Tuesdays and the least amount of Saturdays.

```{r}
ggplot(combined_df, aes(Rain, Count)) + ggtitle("Number of cyclists and the effect of rain")+ geom_point() + geom_boxplot()
```
More rain leads to less than average number of cyclists.


Simple regression model to estimate number of cyclists and effect of rain on cycle activity 

```{r}
combined_df$year <- as.numeric(combined_df$year)
bike.lm <- lm(Count ~ year + season + dow + Rain, data = combined_df)
summary(bike.lm)
```

There is a highly significant effect of year on cyclists, the number of cyclists increases by 1500 every year, on average. There is also a strong seasonal effect, most notably of winter which decreases the number of cyclists by 2400 compared to the baseline autumn. As expected by the graph, there is also a strong effect of day of the week, most notably Tuesday which, on average, has 1400 cyclists more than a typical Monday.

Based on the model, rain does have a large, and highly significant, effect on people cycling in Auckland: each additional millimeter of rain precipitation per day resulted in 225 people less cycling per day. This can be seen in the graphs as well: On days with higher than usual precipitation (i.e more than 40mm) the count of people cycling is significantly less than what is usually the case in that month.
